A Java Wrapper For Embarrassingly Parallel Problems

### 2013-11-26 paper/paper.tex
More paper revisions

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that
can be easily broken up into identical components that do not need to interact
to find the solution. Several problems in the biological sciences are
embarrassingly parallel but they are difficult to rewrite to use multithreading,
because they are either poorly written or too complex. We have written a
framework, implemented in Java, that will execute the serial program in several
threads. These threads working together help cut down the total runtime of the
program exponentially. The next step is to separate the framework into a
front-end that will run locally and a back-end which receives blocks of work and
does the computations. This back-end should run in the cloud, on a machine that
is dedicated to running embarrassingly parallel tasks only.
\end{abstract}

@@

A computation graph consists of nodes that represent the functions or processes.
The edges of the graph show data being passed from one process to another. If
the graph is fully disconnected, none of the processes need to communicate to
complete the computations. Such a computation is called embarrassingly
parallel.\footnote{The term embarrassingly parallel was coined by Cleve Moler in
his 1986 paper ``Matrix Computation on Distributed Memory Multiprocessors.''} In
a practical situation, there will be some inter-node communication to set up the
problem, and then again to accumulate the results.  Embarrassingly parallel
problems can be easily performed on a cluster of servers or in any kind of
distributed network, due to its lack of dependencies.

@@

@@

TODO: Arrows?
The motivation for this effort was based on research into different forms of the
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and
Desi, are very different in size and color. The Desi is also more resistant to
diseases and are more effective at root nodule symbiosis. The Kabuli has a large
but soft shell, which make it at better food source for humans. Both forms are
cultivated and have been under human selection for several thousand years. We
wanted to use an existing Ka/Ks program (Zhang et al. \cite{kaks}) to quantify
the evolution and selection of these chickpeas. We needed to analyze three
variants, each with thousands of sequences. The Ka/Ks Calculator program took
between five to thirty seconds to process one pair of sequences on a commodity
desktop.

At this rate, we predicted it would take ten days of constant computation to
complete the processing we needed. We neither had the time nor the expertise to
develop a new parallel Ka/Ks program. We were also reluctant to revise the
existing code to to implement the parallel computation because of the danger of
introducing bugs into unfamiliar code and because of the learning curve required
for C++ threading. Instead we wrote, a simple Java program to break up the input
and spawn threads that would run the Ka/Ks Calculator process without having to
make any changes to the code. When al of the threads ended, the contents of the
output files were concatenated together to produce a final output that matches
the format and order that would have been produced by a single invocation of the
Ka/Ks Calculator program.

%TODO: get a better source? Talk about UCLUST and octupus
Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters.
Hierarchical clustering algorithms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matrix. There
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would
benefit from parallelism in this field, it was obvious it would be desirable to
develop a framework for the  parallelization task. \cite{cluster}

%TODO: TALK ABOUT CLOUD PAPER HERE.
Segue. Gunarathne et al.\cite{cloud} have investigated running much larger
embarrassingly parallel biological problems using various cloud services. These
services offer on-demand computational power with MORE.
These services also have built in hooks to various distributed services, such as
storage and databases. One of these technologies is the MapReduce framework,
which allows user to perform distributed computation similar to this framework.
As seen in figure X, Apache Hadoop, an open-source implementation of MapReduce,
breaks down the input data into parts, runs them in parallel with the 'Map'
function and then combines the results with the 'Reduce' function. TWO MORE
SENTENCES.

>>>>

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that
can be easily broken up into identical components that do not need to interact
to find the solution. Many problems in the biological sciences are 
embarrassingly parallel but ofter their authors do not have the expertise to
implement a parallel solution. We have written a framework, implemented in Java,
that will execute the serial program in several threads. These threads working
together help cut down the total runtime of the program significantly. The next
step is to separate the framework into a front-end that will run locally and a
back-end which receives blocks of work and does the computations. This back-end
should run in the cloud, on a machine that is dedicated to running
embarrassingly parallel tasks only.
\end{abstract}

@@

A computation graph consists of nodes that represent the functions or processes.
and edges that represent data being passed. If the graph is fully disconnected,
none of the processes need to communicate to complete the computations. Such a
computation is called embarrassingly parallel\footnote{The term embarrassingly
parallel was coined by Cleve Moler in his 1986 paper ``Matrix Computation on
Distributed Memory Multiprocessors.''}. In a practical situation, there will be
some inter-node communication to set up the problem, and then again to
accumulate the results.  Embarrassingly parallel problems can be easily
performed on a cluster of servers or in any kind of distributed network, due to
the lack of dependencies.

@@

%TODO: Embarrassingly parallel diagram
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}

@@

% TODO: Arrows?
The motivation for this effort was based on research into different forms of the
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and
Desi, are very different in size and color. The Desi is also more resistant to
diseases and are more effective at root nodule symbiosis. The Kabuli has a large
but soft shell, which make it at better food source for humans. Both forms are
cultivated and have been under human selection for several thousand years. We
wanted to use an existing Ka/Ks program (Zhang et al. \cite{kaks}) to quantify
the evolution and selection of these chickpeas. We needed to analyze three
variants, each with thousands of sequences. The Ka/Ks Calculator program took
between five to thirty seconds to process one pair of sequences on a ordinary
machine.

At this rate, we predicted it would take upwards of ten days of constant
computation to complete the processing we needed. We neither had the time nor
the expertise to develop a new parallel Ka/Ks program. We were also reluctant to
revise the existing code to to implement the parallel computation because of the
danger of introducing bugs into unfamiliar code and because of the learning
curve required for C++ threading. Instead we wrote, a simple Java program to
break up the input and spawn threads that would run the Ka/Ks Calculator process
without having to make any changes to the code. When al of the threads ended,
the contents of the output files were concatenated together to produce a final
output that matches the format and order that would have been produced by a
single invocation of the Ka/Ks Calculator program.

%TODO: get a better source? Talk about UCLUST and octupus
% Talk about uclust and octupus and blast, even though there are parallelize
% versions. Rewrite to not go into details about the programs but to talk about
% that there are many programs.
Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters.
Hierarchical clustering algorithms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matrix. There
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would
benefit from parallelism in this field, it was obvious it would be desirable to
develop a framework for the parallelization task. \cite{cluster}

%TODO: Footnotes for days
This framework is designed to run locally but there are other options when more
power is needed. Gunarathne et al.\cite{cloud} have investigated running much
larger embarrassingly parallel biological application using various cloud
services. These services offer on-demand computational power with flexibility.
These services have built in hooks to various distributed services, such as
storage and databases. Amazon has a set of cloud computing services called
Amazon Web Services(AWS) and Microsoft has a similar service named Microsoft
Azure. AWS and Azure are esentially virtual instances run on machines in the
cloud that can be spun up for extra computing power and shut down when not
needed.

Amazon also offers another service called Amazon Elastic MapReduce. This
combines its virtual instance framework with Apache Hadoop. As seen
in figure X, Hadoop breaks down the input data into parts, runs them in parallel
with the 'Map' function and then combines the results with the 'Reduce'
function. Dryad is a framework developed released by Microsoft Research that
allows the end-user to create directed acyclic computation graphs that run on
Windows High Performance Computer(HPC) clusters. Gunarathne et al. compared
the services using several applications to analyze performance, cost and
usability..

%TODO: Hadoop figure
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}
